A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-23 text/formal-defnition.tex
Revise Section 3.

In this section, we give a formal definition of \LMD, including
the syntax, full reduction, and type system.  In addition to full reduction,
in which any redex at any stage can be reduced, we also give staged reduction,
which models execution as programs.

@@

A kind, which is used to classify types, is either $*$, the kind of
proper types (types that terms inhabit), or $\Pi x\colon\tau.K$, the kind
of type operators that takes $x$ as an argument of type $\tau$ and returns a type
of kind $K$.
% Types
% of terms have kind $*$ and dependent types have $\Pi$-kinds.  For
% example, $\lambda x:\I.x$ has type $\Pi x:\I.\I$, which has $*$ kind.
% \red{この段落は短いので型の段落と結合するか?}
% Types
A type is a type-level constant $X$, which is declared in the signature with its kind, a dependent function type $\Pi x:\tau_1.\tau_2$,
an application $\tau\ M$ of a type (operator) (of $\Pi$-kind) to a term, a code type $\TW_\alpha$, or an $\alpha$-closed type $\F\alpha.\tau$.
An application of a type (operator) of $\Pi$-kind to a term is, for example, $\text{Vector}\ 10$
if type-level constant $\text{Vector}$ has the kind $\Pi x:\I.*$.
A code type $\TW_\alpha \tau$ is for a code fragment of a term of type $\tau$.
An $\alpha$-closed type corresponds to a runnable code fragment.

@@

Terms include ordinary (explicitly typed) \(\lambda\)-terms, constants,
whose type is declared in signature $\Sigma$, and the following five forms
related to multi-stage programming:
$\TB_\alpha M$ represents a code fragment; $\TB_\alpha M$ represents escape;
$\Lambda\alpha.M$ is a stage variable abstraction;
$M\ A$ is an application of stage $A$ to a stage abstraction; and
$\%_\alpha M$ is an operator for cross-stage persistence.

@@

We adopt the tradition of \LLF-like systems, where types of constants and kinds of type-level constants are globally declared in a signature $\Sigma$, which
is a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use Boolean in \LMD, $\Sigma$ includes $\B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Type environments are sequences of triples of a variable, its type, and its stage.
We write \(\textit{dom}(\Sigma)\) and \(\textit{dom}(\Gamma)\)
for the set of (type-level) constants declared in \(\Sigma\) and \(\Gamma\), respectively.   As in other multi-stage calculi~\cite{taha2003environment,Tsukada,Hanada2014},
a variable declaration is associated with a stage
so that a variable can be referenced only at the declared stage.
Constants and type-level constants are not associated with stages;
so, we can use them at any stage.
We define well-formed signature and well-formed type environments later.

The variable $x$ is bound in $M$ of $\lambda x:\tau.M$ and in $\tau_2$
of $\Pi x:\tau_1.\tau_2$, as usual; the stage variable $\alpha$ is
bound in $M$ of $\Lambda \alpha.M$ and $\tau$ of $\F\alpha.\tau$.
We write $\FV(M)$ and $\FTV(M)$ for the set of free variables and the set of free stage variables in $M$, respectively.  Similarly, $\FV(\tau)$, $\FTV(\tau)$,
$\FV(K)$, and $\FTV(K)$ are defined.
We sometimes abbreviate $\Pi x:\tau_1.\tau_2$ to $\tau_1 \rightarrow \tau_2$ if
$x$ is not a free variable in $\tau_2$.
% Free variables
We identify $\alpha$-convertible terms and assume the names of bound variables are pairwise distinct.
% \AI{This should be mentioned after free variables are introduced.}
% \AI{What about other binders such as $\Pi$?}

@@

\paragraph{Remark:} Basically, we define \LMD to be an extension of
\LTP with dependent types.  One notable difference is that \LMD has
only one kind of \(\alpha\)-closed types, whereas \LTP has two kinds
of \(\alpha\)-closed types \(\forall\alpha.\tau\) and
\(\forall^\varepsilon\alpha.\tau\).  We have omitted the first kind,
for simplicity, and dropped the superscript $\varepsilon$.  It would
not be difficult to recover the distinction to show properties related
to program residualization~\cite{Hanada2014}.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define six kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are
ordinary capture-avoiding substitution of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively,
and we omit their definitions here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are
substitutions of stage $A$ for stage variable $\alpha$ in
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$, respectively.
We also define substitutions of a stage or a term for variables in type environment $\G$.
We show representative cases below.
%
{%\small
\begin{align*}
    (\lambda x:\tau.M)[\alpha \mapsto A] & = \lambda x:(\tau[\alpha \mapsto A]).(M[\alpha \mapsto A])                                  \\
    (M\ B)[\alpha \mapsto A]             & = (M[\alpha \mapsto A])\ B[\alpha\mapsto A]                                                 \\
    (\TB_\beta M)[\alpha \mapsto A]      & = \TB_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                          \\
    (\TBL_\beta M)[\alpha \mapsto A]     & = \TBL_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                         \\
    (\%_\beta M)[\alpha \mapsto A]       & = \%_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                           \\
    (\beta B)[\alpha \mapsto A]          & = \beta (B[\alpha\mapsto A])                               & (\text{if } \alpha \neq \beta) \\
    (\beta B)[\alpha \mapsto A]          & = A (B[\alpha\mapsto A])                                   & (\text{if } \alpha = \beta)
\end{align*}
}
Here, $\TB_{\alpha_1\cdots\alpha_n} M$,
$\TBL_{\alpha_1\cdots\alpha_n} M$, and $\%_{\alpha_1\cdots\alpha_n} M$
$(n \geq 0)$ stand for $\TB_{\alpha_1} \cdots \TB_{\alpha_n} M$,
$\TBL_{\alpha_1}\cdots \TBL_{\alpha_n} M$, and
$\%_{\alpha_1}\cdots \%_{\alpha_n} M$, respectively.  In particular,
$\TB_{\varepsilon} M = \TBL_{\varepsilon} M = \%_{\varepsilon} M = M$.

@@

In this section, we define the type system of \LMD.
It consists of eight judgment forms for signature well-formedness, type environment well-formedness, kind well-formedness, kinding, typing, term equivalence, type equivalence, and kind equivalence.
We list the judgments forms in Figure~\ref{fig:LMD-six-judgments}.
They are all defined in a mutual recursive manner.  We will discuss
each judgment below.

@@

To add declarations to a signature, the kind/type of a (type-level)
constant has to be well-formed at stage \(\varepsilon\) so that it is
used at any stage.

In what follows, well-formedness is not explicitly mentioned but
we assume that all signatures and type environments are well-formed.

@@

The rules for kind wellformedness and kinding is straightforward
adaptation from \LLF and \LTP, except for the rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit.

@@

% \TConv
As in standard dependent type systems, \TConv{} allows us to replace the type of a term with an equivalent one.
For example, when we use a vector type with its size ($\textrm{Vector}\ n$),
a value of type $\textrm{Vector}\ (4+1)$ can also has type $\textrm{Vector}\ 5$ because of \TConv{}.

% Typing rules for a multi-stage calculus
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are rules for a multi-stage calculus.
\TTB and \TTBL are the same with \LTP as we explained in Section \ref{sec:informal-overview}.
The rule \TGen{} for stage abstraction is straightforward.
The condition $\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensures that the scope of $\alpha$ is in $M$.
The rule \TIns{} is for applications of stages to stage abstractions.
\TIns{} is mainly used when we run a code value or rename a stage variable in the body of an abstraction.
The rule \TCsp{} is the rule for CSP, which means that
if term $M$ is of type $\tau$ at stage $A$, $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$.
Note that CSP is also applied to the type \(\tau\) (although it is implicit) in the conclusion.  Thanks to implicit CSP, the typing rule is the same as \LTP.

@@

% The term equivalence judgment of the form $\G \V M \E N : \rho @ A$,
% which means that terms $M$ and $N$ are equivalent as terms of type
% $\rho$ at stage $A$ under $\G$, is defined by the rules in
We show the rules for term equivalence are shown in
Figure~\ref{fig:term-equivalence-rules}, omitting
straightforward rules for reflexivity, symmetry, transitivity,
and compatibility.
% Most rules are
% straightforward.  The rules \QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns,
% \QCsp, \QRefl, \QSym, and \QTrans{} make the relation congruence;
The rules \QBeta, \QTBLTB, and \QLambda{} correspond to
$\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction, respectively.

@@

The helper function vadd$_1$ takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn                                \\
	  & \hspace{7mm} = \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.            \\
	  & \hspace{14mm} \textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}                                                                               \\
	  & \hspace{14mm} \textbf{else}\ \TB_\alpha (                                                                                       \\
	  & \hspace{21mm} \textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2))
\end{align*}
(Here, we assume that the type system can know whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)

The main function vadd below takes the length $n$ and returns
($\beta$-closed) code of a function that takes two vectors and returns
their addition using vadd$_1$.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{7mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta (\text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))) 
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}
If this function is applied to a concrete
(nonnegative) integer, say 5, it returns function code for adding
two vectors of size 5.  The type of vadd\ 5, obtained by
substituting 5 for $n$, is
$\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
\renewcommand{\Vpn}{\text{Vector}\ 5}
If the obtained code is run by applying to \(\varepsilon\),
the type of vadd\ 5\ $\varepsilon$ is
\(\Vpn\to\Vpn\to\Vpn\) as expected.
Without \QPercent, however, it would not be possible
to compose this function code with another piece of code
$v$ of type, say, \(\TW_\gamma \text{Vector }5\) because
\(\text{Vector }5\) and \(\text{Vector }(\%_\beta 5)\) are
syntactically different types.
% : \(\TB_\gamma (\TBL_\gamma (\text{vadd}\ 5\ \gamma)\ v\ v)\).

@@

\renewcommand{\Vpn}{\text{Vector}\ n}
\paragraph{Remark:} By using implicit type-level CSP, the type of vadd
could have been written
\(\Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)\).  In this type,
Vector\ $n$ is given kind at stage \(\varepsilon\) and type-level CSP
implicitly lifts it to stage \(\beta\).  However, if a type-level
constant takes two or more arguments from different stages, term-level
CSP is necessary.  A matrix type (indexed by the numbers of columns
and rows) is such an example.

@@

The reduction given above is full reduction and any redexes---even
under quotation---can be reduced in an arbitrary order.
Following previous work~\cite{Hanada2014},
we introduce (small-step, call-by-value) staged semantics,
where only $\beta$-reduction or $\Lambda$-reduction at stage $\varepsilon$ or the outer-most $\blacklozenge$-reduction are allowed,
modeling an implementation.

@@

Values at $\varepsilon$ stage are a $\lambda$-abstraction, a quoted code,
or a $\Lambda$ abstraction.  The body of a $\lambda$-abstraction can
be any term but the body of $\Lambda$-abstraction has to a value.  It
means that the body of $\Lambda$-abstraction must be evaluated.  The
side condition for $\TBL_\alpha v^{A'}$ means that escapes in a value
can appear only under nested quotations (because an escape under a
single quotation will splice the code value into the surrounding
code).  See Hanada and Igarashi~\cite{Hanada2014} for details.

@@

\begin{definition}[Staged Reduction]
  The staged reduction relation, written $M \longrightarrow_s M'$, is defined by
  the least relation closed under the rules below.
  \begin{align*}
    E^A_\varepsilon [(\lambda x:\tau.M)\ v^\varepsilon] & \longrightarrow_s E^A_\varepsilon[M[x\mapsto v^\varepsilon]]      \\
    E^A_\varepsilon [(\Lambda\alpha.v^\varepsilon)\ A]  & \longrightarrow_s E^A_\varepsilon[v^\varepsilon[\alpha\mapsto A]] \\
    E^A_\alpha [\TBL_\alpha \TB_\alpha v^\alpha]        & \longrightarrow_s E^A_\alpha[v^\alpha]                            \\
  \end{align*}
\end{definition}

>>>>

In this section, we give a formal definition of \LMD, including
the syntax, full reduction, and type system.  In addition to the full reduction,
in which any redex at any stage can be reduced, we also give staged reduction,
which models program execution (at \(\varepsilon\)-stage).

@@

A kind, which is used to classify types, is either $*$, the kind of
proper types (types that terms inhabit), or $\Pi x\colon\tau.K$, the kind
of type operators that takes $x$ as an argument of type $\tau$ and returns a type
of kind $K$.
% Types
% of terms have kind $*$ and dependent types have $\Pi$-kinds.  For
% example, $\lambda x:\I.x$ has type $\Pi x:\I.\I$, which has $*$ kind.
% \red{この段落は短いので型の段落と結合するか?}
% Types
A type is a type-level constant $X$, which is declared in the signature with its kind, a dependent function type $\Pi x:\tau.\sigma$,
an application $\tau\ M$ of a type (operator of $\Pi$-kind) to a term, a code type $\TW_\alpha \tau$, or an $\alpha$-closed type $\F\alpha.\tau$.
An example of an application of a type (operator) of $\Pi$-kind to a term is $\text{Vector}\ 10$; it is well kinded
if, say, type-level constant $\text{Vector}$ has kind $\Pi x:\I.*$.
A code type $\TW_\alpha \tau$ is for a code fragment of a term of type $\tau$.
An $\alpha$-closed type, when used with $\TW_\alpha$, represents runnable code.

@@

Terms include ordinary (explicitly typed) \(\lambda\)-terms, constants,
whose types are declared in signature $\Sigma$, and the following five forms
related to multi-stage programming:
$\TB_\alpha M$ represents a code fragment; $\TBL_\alpha M$ represents escape;
$\Lambda\alpha.M$ is a stage variable abstraction;
$M\ A$ is an application of stage $A$ to a stage abstraction $M$; and
$\%_\alpha M$ is an operator for cross-stage persistence.

@@

We adopt the tradition of \LLF-like systems, where types of constants and kinds of type-level constants are globally declared in a signature $\Sigma$, which
is a sequence of declarations of the form $c:\tau$ and $X::K$.
For example, when we use Boolean in \LMD, $\Sigma$ includes $\B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Type environments are sequences of triples of a variable, its type, and its stage.
We write \(\textit{dom}(\Sigma)\) and \(\textit{dom}(\Gamma)\)
for the set of (type-level) constants and variables declared in \(\Sigma\) and \(\Gamma\), respectively.   As in other multi-stage calculi~\cite{taha2003environment,Tsukada,Hanada2014},
a variable declaration is associated with a stage
so that a variable can be referenced only at the declared stage.
On the contrary, constants and type-level constants are \emph{not} associated with stages;
so, they can appear at any stage.
We define well-formed signature and well-formed type environments later.

The variable $x$ is bound in $M$ of $\lambda x:\tau.M$ and in $\sigma$
of $\Pi x:\tau.\sigma$, as usual; the stage variable $\alpha$ is
bound in $M$ of $\Lambda \alpha.M$ and $\tau$ of $\F\alpha.\tau$.
The notion of free variables is defined in a standard manner.
We write $\FV(M)$ and $\FTV(M)$ for the set of free variables and the set of free stage variables in $M$, respectively.  Similarly, $\FV(\tau)$, $\FTV(\tau)$,
$\FV(K)$, and $\FTV(K)$ are defined.
We sometimes abbreviate $\Pi x:\tau_1.\tau_2$ to $\tau_1 \rightarrow \tau_2$ if
$x$ is not a free variable of $\tau_2$.
% Free variables
We identify $\alpha$-convertible terms and assume the names of bound variables are pairwise distinct.
% \AI{This should be mentioned after free variables are introduced.}
% \AI{What about other binders such as $\Pi$?}

@@

\paragraph{Remark:} Basically, we define \LMD to be an extension of
\LTP with dependent types.  One notable difference is that \LMD has
only one kind of \(\alpha\)-closed types, whereas \LTP has two kinds
of \(\alpha\)-closed types \(\forall\alpha.\tau\) and
\(\forall^\varepsilon\alpha.\tau\).  We have omitted the first kind,
for simplicity, and dropped the superscript $\varepsilon$ from the second.  It would
not be difficult to recover the distinction to show properties related
to program residualization~\cite{Hanada2014}.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define two kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are
ordinary capture-avoiding substitution of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively,
and we omit their definitions here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are
substitutions of stage $A$ for stage variable $\alpha$ in
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$, respectively.
We show representative cases below.
%
{%\small
\begin{align*}
    (\lambda x:\tau.M)[\alpha \mapsto A] & = \lambda x:(\tau[\alpha \mapsto A]).(M[\alpha \mapsto A])                                  \\
    (M\ B)[\alpha \mapsto A]             & = (M[\alpha \mapsto A])\ B[\alpha\mapsto A]                                                 \\
    (\TB_\beta M)[\alpha \mapsto A]      & = \TB_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                          \\
    (\TBL_\beta M)[\alpha \mapsto A]     & = \TBL_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                         \\
    (\%_\beta M)[\alpha \mapsto A]       & = \%_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                           \\
    (\beta B)[\alpha \mapsto A]          & = \beta (B[\alpha\mapsto A])                               & (\text{if } \alpha \neq \beta) \\
    (\beta B)[\alpha \mapsto A]          & = A (B[\alpha\mapsto A])                                   & (\text{if } \alpha = \beta)
\end{align*}
}
Here, $\TB_{\alpha_1\cdots\alpha_n} M$,
$\TBL_{\alpha_1\cdots\alpha_n} M$, and $\%_{\alpha_1\cdots\alpha_n} M$
$(n \geq 0)$ stand for $\TB_{\alpha_1} \cdots \TB_{\alpha_n} M$,
$\TBL_{\alpha_1}\cdots \TBL_{\alpha_n} M$, and
$\%_{\alpha_1}\cdots \%_{\alpha_n} M$, respectively.  In particular,
$\TB_{\varepsilon} M = \TBL_{\varepsilon} M = \%_{\varepsilon} M = M$.
We also define substitutions of a stage or a term for variables in type environment $\G$.

@@

In this section, we define the type system of \LMD.
It consists of eight judgment forms for signature well-formedness, type environment well-formedness, kind well-formedness, kinding, typing, kind equivalence, type equivalence, and term equivalence.
We list the judgments forms in Figure~\ref{fig:LMD-six-judgments}.
They are all defined in a mutual recursive manner.  We will discuss
each judgment below.

@@

To add declarations to a signature, the kind/type of a (type-level)
constant has to be well-formed at stage \(\varepsilon\) so that it is
used at any stage.  In what follows, well-formedness is not explicitly
mentioned but we assume that all signatures and type environments are
well-formed.

@@

The rules for kind wellformedness and kinding is a straightforward
adaptation from \LLF and \LTP, except for the following rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit.

@@

% \TConv
As in standard dependent type systems, \TConv{} allows us to replace the type of a term with an equivalent one.
For example, assuming integers and arithmetic, a value of type $\textrm{Vector}\ (4+1)$ can also has type $\textrm{Vector}\ 5$ because of \TConv{}.

% Typing rules for a multi-stage calculus
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are constructs for multi-stage progarmming.
\TTB{} and \TTBL{} are the same as \LTP, as we explained in Section \ref{sec:informal-overview}.
The rule \TGen{} for stage abstraction is straightforward.
The condition $\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensures that the scope of $\alpha$ is in $M$.
The rule \TIns{} is for applications of stages to stage abstractions.
% \TIns{} is mainly used when we run a code value or rename a stage variable in the body of an abstraction.
The rule \TCsp{} is for CSP, which means that,
if term $M$ is of type $\tau$ at stage $A$, then $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$.
Note that CSP is also applied to the type \(\tau\) (although it is implicit) in the conclusion.  Thanks to implicit CSP, the typing rule is the same as \LTP.

@@

% The term equivalence judgment of the form $\G \V M \E N : \rho @ A$,
% which means that terms $M$ and $N$ are equivalent as terms of type
% $\rho$ at stage $A$ under $\G$, is defined by the rules in
We show the rules for term equivalence  in
Figure~\ref{fig:term-equivalence-rules}, omitting
straightforward rules for reflexivity, symmetry, transitivity,
and compatibility.
% Most rules are
% straightforward.  The rules \QAbs, \QApp, \QTB, \QTBL, \QGen, \QIns,
% \QCsp, \QRefl, \QSym, and \QTrans{} make the relation congruence;
The rules \QBeta, \QTBLTB, and \QLambda{} correspond to
$\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction, respectively.

@@

The helper function vadd$_1$ takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{tabbing}
	  $\textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn$                                \\
	  \hspace{6mm} \= $= \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.$            \\
	  \> \hspace{6mm} \= $\textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}$ \\
	  \>\> $\textbf{else}\ \TB_\alpha ($ \= $\textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}$ \\
	  \>\>\> $\textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}$ \\
          \>\>\> $\text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2))$
\end{tabbing}
(Here, we assume that the type system can know whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)

The main function vadd below takes the length $n$ and returns
($\beta$-closed) code of a function that takes two vectors and returns
their addition using vadd$_1$.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{6mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta (\text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))) 
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}%
If this function is applied to a
(nonnegative) integer constant, say 5, it returns function code for adding
two vectors of size 5.  The type of vadd\ 5, obtained by
substituting 5 for $n$, is
$\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
\renewcommand{\Vpn}{\text{Vector}\ 5}
If the obtained code is run by applying to \(\varepsilon\),
the type of vadd\ 5\ $\varepsilon$ is
\(\Vpn\to\Vpn\to\Vpn\) as expected.
Without \QPercent, however, it would not be possible
to compose this function code with another piece of code
$v$ of type, say, \(\TW_\gamma \text{Vector }5\) because
\(\text{Vector }5\) and \(\text{Vector }(\%_\beta 5)\) are
syntactically different types.
% : \(\TB_\gamma (\TBL_\gamma (\text{vadd}\ 5\ \gamma)\ v\ v)\).

@@

\renewcommand{\Vpn}{\text{Vector}\ n}
\paragraph{Remark:} By using implicit type-level CSP, the type of vadd
could have been written
\(\Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)\).  In this type,
Vector\ $n$ is given kind at stage \(\varepsilon\) and type-level CSP
implicitly lifts it to stage \(\beta\).  However, if a type-level
constant takes two or more arguments from different stages, term-level
CSP is necessary.  A matrix type (indexed by the numbers of columns
and rows) would be such an example.

@@

The reduction given above is full reduction and any redexes---even
under $\TB_\alpha$---can be reduced in an arbitrary order.
Following previous work~\cite{Hanada2014},
we introduce (small-step, call-by-value) staged semantics,
where only $\beta$-reduction or $\Lambda$-reduction at stage $\varepsilon$ or the outer-most $\blacklozenge$-reduction are allowed,
modeling an implementation.

@@

Values at $\varepsilon$ stage are a $\lambda$-abstraction, a quoted code,
or a $\Lambda$-abstraction.  The body of a $\lambda$-abstraction can
be any term but the body of $\Lambda$-abstraction has to a value.  It
means that the body of $\Lambda$-abstraction must be evaluated.  The
side condition for $\TBL_\alpha v^{A'}$ means that escapes in a value
can appear only under nested quotations (because an escape under a
single quotation will splice the code value into the surrounding
code).  See Hanada and Igarashi~\cite{Hanada2014} for details.

@@

\begin{definition}[Staged Reduction]
  The staged reduction relation, written $M \longrightarrow_s N$, is defined by
  the least relation closed under the rules below.
  \begin{align*}
    E^A_\varepsilon [(\lambda x:\tau.M)\ v^\varepsilon] & \longrightarrow_s E^A_\varepsilon[M[x\mapsto v^\varepsilon]]      \\
    E^A_\varepsilon [(\Lambda\alpha.v^\varepsilon)\ A]  & \longrightarrow_s E^A_\varepsilon[v^\varepsilon[\alpha\mapsto A]] \\
    E^A_\alpha [\TBL_\alpha \TB_\alpha v^\alpha]        & \longrightarrow_s E^A_\alpha[v^\alpha]                            \\
  \end{align*}
\end{definition}
